---
title: "Predicting the Winner of 2021 D1 NCAA Basketball Tournament"
author: "David Nemirovsky"
output: pdf_document
date: 3/22/21
--- 

```{r setup, include = FALSE}
library(tidyverse)
library(caret)
library(splines)
library(mgcv)
library(pdp)
library(earth)
library(pROC)
library(vip)
library(AppliedPredictiveModeling)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = .6,
  out.width = "95%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

`%nin%` = Negate(`%in%`)

set.seed(37564)
```

## **EDA**

+ Build data frames for exploratory data analysis and model construction:

```{r dataframes, message = F, warning = F}
season_df = 
  read_csv("./data/MRegularSeasonDetailedResults.csv") %>% 
  rowid_to_column("game_id") %>% 
  relocate(WLoc:NumOT, .after = DayNum) %>% 
  pivot_longer(
    WTeamID:LPF,
    names_to = "stat",
    values_to = "count"
  ) %>% 
  mutate(outcome = case_when(
    str_detect(stat, "^W") ~ "win", 
    str_detect(stat, "^L") ~ "loss"
  )) %>% 
  mutate(stat = substr(stat, 2, nchar(stat))) %>% 
  pivot_wider(
    names_from = stat,
    values_from = count
  ) %>% 
  mutate(TeamID = as.factor(TeamID)) %>% 
  filter(Season > 2014) %>% 
  mutate(
    Season = as.factor(Season), 
    outcome = as.factor(outcome), 
    fg_pct = FGM/FGA, 
    fg3_pct = FGM3/FGA3, 
    ft_pct = FTM/FTA) %>% 
  select(-(FGM:FTA)) %>% 
  unnest()

opp_df = 
  season_df %>% 
  group_by(game_id) %>% 
  arrange(outcome, .by_group = T) %>% 
  ungroup() %>% 
  select(OR:ft_pct) %>% 
  rename_all(~paste0("opp", .x))

train_df = 
  season_df %>% 
  cbind(opp_df) %>% 
  as_tibble() %>% 
  mutate(
    OR_diff = OR - oppOR,
    DR_diff = DR - oppDR,
    Ast_diff = Ast - oppAst,
    TO_diff = TO - oppTO,
    Stl_diff = Stl - oppStl,
    Blk_diff = Blk - oppBlk,
    PF_diff = PF - oppPF,
    FGpct_diff = fg_pct - oppfg_pct,
    FG3pct_diff = fg3_pct - oppfg3_pct,
    FTpct_diff = ft_pct - oppft_pct
  ) %>% 
  select(Season, outcome, TeamID, OR_diff:FTpct_diff)

tourney_df = 
  read_csv("./data/MNCAATourneyDetailedResults.csv") %>% 
  rowid_to_column("game_id") %>% 
  relocate(WLoc:NumOT, .after = DayNum) %>% 
  pivot_longer(
    WTeamID:LPF,
    names_to = "stat",
    values_to = "count"
  ) %>% 
  mutate(outcome = case_when(
    str_detect(stat, "^W") ~ "win", 
    str_detect(stat, "^L") ~ "loss"
  )) %>% 
  mutate(stat = substr(stat, 2, nchar(stat))) %>% 
  pivot_wider(
    names_from = stat,
    values_from = count
  ) %>% 
  mutate(TeamID = as.factor(TeamID)) %>% 
  filter(Season > 2014) %>% 
  mutate(
    Season = as.factor(Season), 
    outcome = as.factor(outcome), 
    fg_pct = FGM/FGA, 
    fg3_pct = FGM3/FGA3, 
    ft_pct = FTM/FTA) %>% 
  select(-(FGM:FTA)) %>% 
  unnest()

opp_tourney = 
  tourney_df %>% 
  group_by(game_id) %>% 
  arrange(outcome, .by_group = T) %>% 
  ungroup() %>% 
  select(OR:ft_pct) %>% 
  rename_all(~paste0("opp", .x))

test_df = 
  tourney_df %>% 
  cbind(opp_tourney) %>% 
  as_tibble() %>% 
  mutate(
    OR_diff = OR - oppOR,
    DR_diff = DR - oppDR,
    Ast_diff = Ast - oppAst,
    TO_diff = TO - oppTO,
    Stl_diff = Stl - oppStl,
    Blk_diff = Blk - oppBlk,
    PF_diff = PF - oppPF,
    FGpct_diff = fg_pct - oppfg_pct,
    FG3pct_diff = fg3_pct - oppfg3_pct,
    FTpct_diff = ft_pct - oppft_pct
  ) %>% 
  select(Season, outcome, TeamID, OR_diff:FTpct_diff)

total_games_df = 
  season_df %>% 
  group_by(Season, TeamID) %>% 
  summarize(n_games = n())

win_pct_df = 
  season_df %>% 
  group_by(Season, TeamID) %>% 
  filter(outcome == "win") %>% 
  summarize(n_wins = n()) %>% 
  left_join(total_games_df) %>% 
  mutate(win_pct = n_wins/n_games)

regstats_df = 
  season_df %>% 
  select(Season, TeamID, OR:ft_pct) %>% 
  group_by(Season, TeamID) %>% 
  summarize_at(vars(OR:ft_pct), ~mean(.x))

regszn_df = 
  win_pct_df %>% 
  left_join(regstats_df) %>% 
  drop_na() %>% 
  group_split(Season)
```

+ Scatter plots of win probability across different seasonal stats:

```{r eda1}
x = model.matrix(win_pct ~ ., regszn_df[[1]] %>% dplyr::select(-(Season:n_games)))[ ,-1]
y = regszn_df[[1]]$win_pct

theme1 = trellis.par.get()
theme1$plot.symbol$col = rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch = 16
theme1$plot.line$col = rgb(.8, .1, .1, 1)
theme1$plot.line$lwd = 2
theme1$strip.background$col = rgb(.0, .2, .6, .2)
trellis.par.set(theme1)
featurePlot(x, y, plot = "scatter", labels = c("", "Win %"), type = c("p"), layout = c(5, 2))
```

+ Distributions of game stat differences, grouped by outcome:

```{r eda2}
featurePlot(x = train_df[, 4:13], 
            y = train_df$outcome,
            scales = list(x = list(relation = "free"), 
                          y = list(relation = "free")),
            plot = "box", 
            auto.key = list(columns = 2))
```

+ Test three different models (simple logistic regression, MARS, and KNN) using 10-fold cross-validation:

```{r cv, message = F, warning = F}
set.seed(37564)

ctrl = trainControl(method = "cv", summaryFunction = twoClassSummary, classProbs = T, number = 10)

mods_df = 
  tibble(Year = c(2015:2020)) %>% 
  mutate(train = map(.x = Year, ~filter(train_df, Season == .x)), 
         test = map(.x = Year, ~filter(test_df, Season == .x))) %>% 
  mutate(mod_glm = map(.x = train, ~train(outcome ~ . -Season -TeamID,
                                       na.action = na.exclude, 
                                       data = .x, 
                                       method = "glm", 
                                       family = "binomial", 
                                       metric = "ROC", 
                                       trControl = ctrl)), 
         mod_mars = map(.x = train, ~train(outcome ~ . -Season -TeamID,
                                       na.action = na.exclude, 
                                       data = .x, 
                                       method = "earth",
                                       tuneGrid = expand.grid(degree = 1, nprune = 13:20), 
                                       metric = "ROC", 
                                       trControl = ctrl)), 
         mod_knn = map(.x = train, ~train(outcome ~ . -Season -TeamID,
                                       na.action = na.exclude, 
                                       data = .x, 
                                       method = "knn",
                                       metric = "ROC", 
                                       preProcess = c("center","scale"),
                                       tuneGrid = data.frame(k = seq(30, 80, by = 5)),
                                       trControl = ctrl)))

plot_knn = map(.x = c(1:6), ~ggplot(mods_df$mod_knn[[.x]], highlight = T) + theme(plot.title = element_text(hjust = 0.5)))

(plot_knn[[1]] + labs(title = "2015")) + (plot_knn[[2]] + labs(title = "2016")) + (plot_knn[[3]] + labs(title = "2017")) + (plot_knn[[4]] + labs(title = "2018")) + (plot_knn[[5]] + labs(title = "2019")) + (plot_knn[[6]] + labs(title = "2020"))

plot_mars = map(.x = c(1:6), ~ggplot(mods_df$mod_mars[[.x]], highlight = T) + theme(plot.title = element_text(hjust = 0.5)))

(plot_mars[[1]] + labs(title = "2015")) + (plot_mars[[2]] + labs(title = "2016")) + (plot_mars[[3]] + labs(title = "2017")) + (plot_mars[[4]] + labs(title = "2018")) + (plot_mars[[5]] + labs(title = "2019")) + (plot_mars[[6]] + labs(title = "2020"))

res = map(.x = c(1:6), ~resamples(list(GLM = mods_df$mod_glm[[.x]], MARS = mods_df$mod_mars[[.x]], KNN = mods_df$mod_knn[[.x]])))
```

```{r bwplots, figures-side, fig.show = "hold", out.width = "33%"}
bwplot(res[[1]], metric = "ROC", main = "2015")
bwplot(res[[2]], metric = "ROC", main = "2016")
bwplot(res[[3]], metric = "ROC", main = "2017")
bwplot(res[[4]], metric = "ROC", main = "2018")
bwplot(res[[5]], metric = "ROC", main = "2019")
bwplot(res[[6]], metric = "ROC", main = "2020")
```

+ 10-fold CV determined that the GLM using logistic regression is the best model to use.

+ Now, we test the GLM model for each season, excluding 2020 since that will be used for the Tournament predictions.

```{r pred, message = F, warning = F}
pred_df = 
  mods_df %>% 
  filter(Year < 2020) %>% 
  dplyr::select(-mod_mars, -mod_knn) %>% 
  mutate(pred = map2(.x = mod_glm, .y = test, ~predict(.x, .y, type = "prob")[, 2])) %>% 
  unnest(test, pred) %>% 
  dplyr::select(Season, outcome, pred)

plot(roc(pred_df$outcome, pred_df$pred), legacy.axes = TRUE, print.auc = TRUE)
```

+ The AUC of the ROC using the logisitic regression model is **`r round(roc(pred_df$outcome, pred_df$pred)$auc[1], 3)`**, suggesting that this model does a great job of predicting outcomes.

+ Now, let's create a function that will predict the outcome of a matchup between two teams in the tournament, using the GLM logistic regression model:

```{r matchup fct, message = F}
team_df = 
  read_csv("./data/MTeams.csv") %>% 
  mutate(TeamID = as.factor(TeamID), 
         TeamName = as.factor(TeamName))

reg2020 = 
  regszn_df[[6]] %>% 
  left_join(team_df) %>% 
  select(Season, TeamID, TeamName, OR:ft_pct)

model = mods_df$mod_glm[[6]]
summary(model)

matchup = function(Team1, Team2) {
  
  t1_df = 
    reg2020 %>% 
    filter(TeamID %in% c(Team1, Team2))

  t2_df = 
    reg2020 %>% 
    filter(TeamID %in% c(Team1, Team2)) %>% 
    arrange(desc(TeamID)) %>% 
    select(OR:ft_pct) %>% 
    rename_all(~paste0("opp", .x))
  
  test2020 = 
    t1_df %>% 
    cbind(t2_df) %>% 
    as_tibble() %>%
    mutate(
      OR_diff = OR - oppOR,
      DR_diff = DR - oppDR,
      Ast_diff = Ast - oppAst,
      TO_diff = TO - oppTO,
      Stl_diff = Stl - oppStl,
      Blk_diff = Blk - oppBlk,
      PF_diff = PF - oppPF,
      FGpct_diff = fg_pct - oppfg_pct,
      FG3pct_diff = fg3_pct - oppfg3_pct,
      FTpct_diff = ft_pct - oppft_pct
    ) %>% 
    select(Season, TeamID, OR_diff:FTpct_diff)
  
  win_prob = predict(model, test2020, type = "prob")[1, 2]
  
 if (Team1 < Team2) {if (win_prob > 0.5) {
    print(Team1) 
  } else {
    print(Team2)
  }} else {if (win_prob > 0.5) {
    print(Team2) 
  } else {
    print(Team1)
  }}
  
}
```

+ For the first four play-in games:

```{r first four, eval = F}
reg2020 %>% 
  filter(TeamName %in% 
           c("Norfolk St", "Wichita St", "TX Southern", "UCLA", "Michigan St", "Drake", "Appalachian St", "Mt St Mary's")) %>% 
  select(TeamName, TeamID)

matchup(1111, 1313)
matchup(1111, 1313)
matchup(1111, 1313)
matchup(1111, 1313)
```

+ Now, to create the 4 regions and seed all the teams:

```{r seed, message = F}
seed2020 = 
  read_csv("./data/MNCAATourneySeeds.csv") %>% 
  filter(Season == "2021") %>% 
  select(-Season) %>% 
  mutate(TeamID = as.factor(TeamID)) %>% 
  left_join(reg2020) %>% 
  mutate(Region = as.factor(case_when(
    str_detect(Seed, "^W") ~ "East", 
    str_detect(Seed, "^X") ~ "West", 
    str_detect(Seed, "^Y") ~ "Midwest", 
    str_detect(Seed, "^Z") ~ "South"))) %>% 
  separate(Seed, c(NA, "Seed"), sep = "[W-Z]") %>% 
  filter(TeamName %nin% c("Norfolk St", "Wichita St", "TX Southern", "UCLA")) %>%
  mutate(Seed = as.numeric(str_remove(Seed, "[ab]"))) %>% 
  mutate(TeamID = as.character(TeamID), 
         TeamID = as.numeric(TeamID))

East = 
  seed2020 %>% 
  filter(Region == "East")
West = 
  seed2020 %>% 
  filter(Region == "West")
Midwest = 
  seed2020 %>% 
  filter(Region == "Midwest")
South = 
  seed2020 %>% 
  filter(Region == "South")
```

+ Simulate tournament by regions, using self-created `region_sim()` function:

``` {r region sim}
region_sim = function(region) {
  
  r1 = region %>% 
    filter(Seed %in% c(1, 16)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)
  
  r2 = region %>% 
    filter(Seed %in% c(2, 15)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)
  
  r3 = region %>% 
    filter(Seed %in% c(3, 14)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)
  
  r4 = region %>% 
    filter(Seed %in% c(4, 13)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)
  
  r5 = region %>% 
    filter(Seed %in% c(5, 12)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)
  
  r6 = region %>% 
    filter(Seed %in% c(6, 11)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)
  
  r7 = region %>% 
    filter(Seed %in% c(7, 10)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)
  
  r8 = region %>% 
    filter(Seed %in% c(8, 9)) %>% 
    arrange(TeamID) %>% 
    pull(TeamID)

  res1 = matchup(r1[1], r1[2])
  res2 = matchup(r2[1], r2[2])
  res3 = matchup(r3[1], r3[2])
  res4 = matchup(r4[1], r4[2])
  res5 = matchup(r5[1], r5[2])
  res6 = matchup(r6[1], r6[2])
  res7 = matchup(r7[1], r7[2])
  res8 = matchup(r8[1], r8[2])
  
  quart1 = matchup(res1, res8)
  quart2 = matchup(res2, res7)
  quart3 = matchup(res3, res6)
  quart4 = matchup(res4, res5)
  
  sem1 = matchup(quart1, quart4)
  sem2 = matchup(quart2, quart3)
  
  fin = matchup(sem1, sem2)
  
  c(res1, res2, res3, res4, res5, res6, res7, res8, quart1, quart2, quart3, quart4, sem1, sem2, fin) %>% 
    as_tibble() %>% 
    rename(TeamID = value) %>% 
    mutate(TeamID = as.factor(TeamID)) %>% 
    left_join(mutate(seed2020, TeamID = as.factor(TeamID))) %>% 
    dplyr::select(TeamID, Seed, TeamName)

}

region_sim(East)
region_sim(West)
region_sim(Midwest)
region_sim(South)
```

+ Finish tournament by simulating the Final Four, using self-created `f4_sim()` function:

```{r final four sim}
f4_sim = function(regE, regW, regMW, regS){
  
  winEast = region_sim(regE)[[15, 1]] %>% as.character() %>% as.numeric()
  winWest = region_sim(regW)[[15, 1]] %>% as.character() %>% as.numeric()
  winMidwest = region_sim(regMW)[[15, 1]] %>% as.character() %>% as.numeric()
  winSouth = region_sim(regS)[[15, 1]] %>% as.character() %>% as.numeric()
  
  nat_sem1 = matchup(winEast, winWest)
  nat_sem2 = matchup(winMidwest, winSouth)
  
  nat_fin = matchup(nat_sem1, nat_sem2)
  
  c(nat_sem1, nat_sem2, nat_fin) %>% 
    as_tibble() %>% 
    rename(TeamID = value) %>% 
    mutate(TeamID = as.factor(TeamID)) %>% 
    left_join(mutate(seed2020, TeamID = as.factor(TeamID))) %>% 
    dplyr::select(TeamID, Seed, TeamName)

}

f4_sim(East, West, Midwest, South)
```
